---
title: "make Neff statistics per each replicate"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
---


Compute \(N_{\text{eff}}\) statistics for each stochastic-map replicate, summarize results with the mean and 95% confidence interval (2.5–97.5% quantiles) across 15 body regions, and save the output to: `data_out/neff-per-replicate.RDS`

```{r setup, include=F}

library(ape)
library(phytools)
library(ontophylo)
library(dplyr)
library(parallel)
library(viridis)
library(ggplot2)
library(abind)
library(dplyr)
library(purrr)
source("R/utils-ST.R")
source('R/region_class.R')


knitr::opts_chunk$set(
  echo = TRUE,       # show code by default
  warning = FALSE,
  message = FALSE
)
```


# Read data

```{r}

# read replicates
tree <-readRDS("data/hym_tree.RDS")
rates.multi <- readRDS("data_out/br_rates_all.RDS")

class(rates.multi)
length(rates.multi)
dim(rates.multi$cranium)

# read mean rates
rates <-readRDS("data_out/br_rates_all_ind.RDS")
rates <- do.call(cbind,rates)
# drop larva
rates <- rates[,-16]
class(rates)
dim(rates)
```
# Reshape as list of 1000 tables 152 x 15

```{r}
# names of body regions
regions <- names(rates.multi)
n_reps  <- ncol(rates.multi[[1]])   # 1000
n_branches <- nrow(rates.multi[[1]]) # 152

# build new list: one element per replicate
replicate_list <- lapply(1:n_reps, function(j) {
  # extract column j from each body region and bind
  mat <- sapply(rates.multi, function(region_mat) region_mat[, j])
  # ensure it's a matrix with rows = branches, cols = body regions
  mat <- matrix(mat, nrow = n_branches, ncol = length(regions))
  colnames(mat) <- regions
  mat
})

# drop larva
replicate_list <- lapply(replicate_list, function(x) x[, -16, drop = FALSE])

# check dims
length(replicate_list)   # 1000
dim(replicate_list[[1]]) # 152 x 15

# check if rates.multi==rates
col_means <- Reduce("+", replicate_list) / length(replicate_list)
diff <- col_means-rates
max(abs(diff))
```

# Calculate Neff for each replicate

May take a while

```{r}
n_eff_list <- map(replicate_list, function(mat) {
  enr <- get_enrichment_mask(mat, lower_q = 0.05, upper_q = 0.95,
                             direction = "over", return_type = "binary")
  out <- plot_enrichment_entropy(tree, enr, type = "Neff_scaled",
                                 n_points = 1000)$data
  out
})

n_eff_df <- bind_rows(n_eff_list, .id = "replicate")
head(n_eff_df)
```


# Reshape (per replicate Neff as list)

```{r}
summary_df <- n_eff_df %>%
  group_by(time) %>%
  summarise(
    mean_entropy = mean(entropy, na.rm = TRUE),
    lower = quantile(entropy, 0.025, na.rm = TRUE),
    median = quantile(entropy, 0.5, na.rm = TRUE),
    upper = quantile(entropy, 0.975, na.rm = TRUE)
  )
head(summary_df)
```



# Plot median with empirical 95% confidence interval

Alternatively: quantile interval or uncertainty interval

```{r}
ggplot() +
  geom_ribbon(data = summary_df, aes(x = time, ymin = lower, ymax = upper),
              fill = "blue", alpha = 0.2) +
  scale_x_reverse(breaks = seq(0, 280, by = 50)) +
  geom_line(data = summary_df, aes(x = time, y = median),
            color = "black", size = 1) +
  theme_minimal()
```
# Plot mean with empirical 95% confidence interval

```{r}
ggplot() +
  geom_ribbon(data = summary_df, aes(x = time, ymin = lower, ymax = upper),
              fill = "blue", alpha = 0.2) +
  scale_x_reverse(breaks = seq(0, 280, by = 50)) +
  geom_line(data = summary_df, aes(x = time, y = mean_entropy),
            color = "black", size = 1) +
  theme_minimal()
```


# Save

```{r}
saveRDS(summary_df, file='data_out/neff-per-replicate.RDS')
```

