---
title: "Plot Enrichment of Accelerated Rates as Proportion"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
---



```{r setup, include=F}

library(ape)
library(phytools)
library(ontophylo)
library(dplyr)
library(parallel)
library(viridis)
library(ggplot2)
library(deeptime)
library(mgcv)
source("R/utils-ST.R")
source('R/region_class.R')


knitr::opts_chunk$set(
  echo = TRUE,       # show code by default
  warning = FALSE,
  message = FALSE
)
```


## Read and Make Data

```{r}

tree <-readRDS("data/hym_tree.RDS")

# read mean rates
rates <-readRDS("data_out/br_rates_all_ind.RDS")
rates <- do.call(cbind,rates)
# drop larva
rates <- rates[,-16]
class(rates)
dim(rates)

# tot states per BR
load('data/char_num.RDA')
tot_states <- lapply(anat_ent_state, function(x) sum(get_len(x)) ) %>% unlist
#tot_states

# rates_norm <- apply(rates, 1, function(x) x/tot_states) %>% t
# head(rates_norm)
# head(rates)

# calculate full disntces = rate x time
# dist_rates_norm <-  apply(rates_norm, 2, function(x) x*tree$edge.length)

```





## Tip overrepresentation

```{r}

enr.masked <- get_enrichment_mask(rates, lower_q = 0.05, upper_q = 0.95, direction = "over", return_type = "masked")

tip.enr <- get_tip_enrichment(tree, enr.masked)
#tip.enr[tip.enr>0] %>% hist()
hist(tip.enr)
max(tip.enr)
min(tip.enr)
min(tip.enr[tip.enr > 0])
```

### Claculte Enrichment proportion

```{r}
tip.total.changes <- get_tip_enrichment(tree, rates)
tip.total.changes_log <- log(tip.total.changes)
tip.prop.enriched <- tip.enr/tip.total.changes
# tip.enr <-tip.prop
hist(tip.total.changes)
hist(tip.total.changes_log)
```

### Stat for Proportion

#### All tree

```{r}
colMeans(tip.prop.enriched)
mean(tip.prop.enriched)
sd(tip.prop.enriched)


```

#### Time proportion

```{r}
enr.masked.binary <- get_enrichment_mask(rates, lower_q = 0.05, upper_q = 0.95, direction = "over", return_type = "binary")

tip.enr.time <- get_tip_enrichment(tree, enr.masked.binary)
enriched_time_prop <- tip.enr.time/280.0259

colMeans(enriched_time_prop)
mean(enriched_time_prop)

```

##### Plot proportion of changes in enriched branches

```{r}
vals <- colMeans(tip.prop.enriched)
n <- length(vals)

file_base <- "figures/tree-enrich/hym-prop-changes-enriched"

pdf(
  file = paste0(file_base, ".pdf"),
  width = 20,
  height = 9.5,
  useDingbats = FALSE
)

op <- par(no.readonly = TRUE)

par(
  mar = c(12, 4, 2, 1),  # bottom, left, top, right
  xpd = NA              # allow labels outside plot region
)

# Background bars
bp <- barplot(
  rep(1, n),
  col = "gray90",
  border = NA,
  names.arg = names(vals),
  las = 2,
  cex.names = 0.7,
  ylim = c(0, 1),
  axes = FALSE,
  ylab = "Proportion"
)

axis(
  side = 2,
  at = c(0, 0.1, 0.5, 1),
  labels = c("0", "0.1", "0.5", "1"),
  las = 1
)

barplot(
  vals,
  col = "#e31a1c",
  border = NA,
  add = TRUE,
  names.arg = FALSE,
  axes = FALSE
)

par(op)
dev.off()
```


##### Plot proportion of time in enriched branches

```{r}
vals <- colMeans(enriched_time_prop)
n <- length(vals)

file_base <- "figures/tree-enrich/hym-prop-time-enriched"

pdf(
  file = paste0(file_base, ".pdf"),
  width = 20,
  height = 9.5,
  useDingbats = FALSE
)

op <- par(no.readonly = TRUE)

par(
  mar = c(12, 4, 2, 1),  # bottom, left, top, right
  xpd = NA              # allow labels outside plot region
)

# Background bars
bp <- barplot(
  rep(1, n),
  col = "gray90",
  border = NA,
  names.arg = names(vals),
  las = 2,
  cex.names = 0.7,
  ylim = c(0, 1),
  axes = FALSE,
  ylab = "Proportion"
)

axis(
  side = 2,
  at = c(0, 0.1, 0.5, 1),
  labels = c("0", "0.1", "0.5", "1"),
  las = 1
)

barplot(
  vals,
  col = "green",
  border = NA,
  add = TRUE,
  names.arg = FALSE,
  axes = FALSE
)

par(op)
dev.off()
```


## Proportion for tip enrichment

### make tb

```{r}
library(tidyverse)

plot_df <- as.data.frame(tip.prop.enriched) |>
  mutate(row = row_number()) |>
  pivot_longer(
    -row,
    names_to = "region",
    values_to = "prop"
  ) |>
  mutate(
    region = factor(region, levels = colnames(tip.prop.enriched)),
    x = as.numeric(region)
  )
```


### plot

```{r}

p <- ggplot(plot_df) +
  # background 100% cells
  geom_rect(
    aes(
      xmin = x - 0.45,
      xmax = x + 0.45,
      ymin = row - 0.45,
      ymax = row + 0.45
    ),
    fill = "grey90",
    colour = "white",
    linewidth = 0.1
  ) +
  # red proportion, left-aligned
  geom_rect(
    aes(
      xmin = x - 0.45,
      xmax = x - 0.45 + 0.9 * prop,
      ymin = row - 0.45,
      ymax = row + 0.45
    ),
    fill = "#e31a1c"
  ) +
  # 50% vertical dashed line in each column
  geom_segment(
    data = data.frame(
      x = seq_along(levels(plot_df$region)),
      x50 = seq_along(levels(plot_df$region)) - 0.45 + 0.9 * 0.5,
      ymin = 0.5,
      ymax = max(plot_df$row) + 0.5
    ),
    aes(
      x = x50,
      xend = x50,
      y = ymin,
      yend = ymax
    ),
    linetype = "dashed",
    colour = "grey30",
    linewidth = 0.3
  ) +
  scale_x_continuous(
    breaks = seq_along(levels(plot_df$region)),
    labels = levels(plot_df$region),
    position = "top"
  ) +
 scale_y_continuous(
    breaks = NULL,
    labels = NULL
  ) +
  labs(
    x = NULL,
    y = NULL
  ) +
  theme_classic() +
  theme(
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),

    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),

    axis.text.x = element_text(
      angle = 45,
      hjust = 0,
      vjust = 0
    )
  )

ggsave(
  "figures/tree-enrich/hym-proportion-enriched-matrix.pdf",
  plot = p,
  width = 10,
  height = 6,
  units = "in",
  device = cairo_pdf
)
```


## Make colors for heatmap

```{r}


# Make palette
pal_info <- list()
pal_info$pal <- gplots::rich.colors(64, palette="temperature", alpha=1.0, rgb=FALSE, plot=F)
pal_info$pal[1:8] <- pal_info$pal[9]
# breaks<- BAMMtools::assignColorBreaks(c(tip.total.changes[tip.total.changes > 0]), spex=NULL, NCOLORS = 64, method="jenks")
breaks<- BAMMtools::assignColorBreaks(c(tip.total.changes_log), spex=NULL, NCOLORS = 64, method="jenks")
pal_info$breaks <- unique(breaks)

# Fun: vals -> colors
map_to_color <- function(vals, pal_info) {
  cols <- pal_info$pal[cut(vals, breaks = pal_info$breaks, include.lowest = TRUE)]
  #cols[vals == log(eps)] <- "#ffffff"  # force exact zeros to white
  #cols[vals == 0] <- "#ffffff"
  #cols[vals == 0] <- "lightgrey"
  
  #cols[vals == 0] <- "aliceblue"  #"skyblue1" "#deebf7"  # this is a special color for 0s
  cols
}
# map_to_color(tip.total.changes[1,1], pal_info)


# TEST PALETTE
# val_min <- min(tip.total.changes[tip.total.changes > 0]) #min(tip.total.changes, na.rm = TRUE)
# val_max <- max(tip.total.changes, na.rm = TRUE)
# vals_x <- seq(val_min , val_max, length.out=100)
# cols_x <- map_to_color(vals_x, pal_info)
# par(mar = c(2, 4, 2, 2))
# plot(vals_x, rep(1, length(vals_x)),
#      col = cols_x, pch = 15, cex = 3,
#      xlab = "Value", ylab = "", yaxt = "n")
```


### Make heatmap legend

```{r}

# tip.total.changes

# if (!dir.exists("plots")) dir.create("plots")
file_base <- "figures/tree-enrich/heat_map-legend"

# png(
#   file = paste0(file_base, ".png"),
#   width = 15, height = 9, units = "in", res = 300
# )

pdf(
  file = paste0(file_base, ".pdf"),
  width = 15, height = 9
)


# ---- Horizontal legend (point-based, aligned with ticks) ----
# val_min <- min(tip.total.changes[tip.total.changes > 0]) 
# val_max <- max(tip.total.changes, na.rm = TRUE)

val_min <- min(tip.total.changes_log) 
val_max <- max(tip.total.changes_log, na.rm = TRUE)

# values for gradient
vals_x <- seq(val_min, val_max, length.out = 200)
cols_x <- map_to_color(vals_x, pal_info)

par(mar = c(4, 4, 2, 2))
plot(vals_x, rep(0.001, length(vals_x)), # rep(0.001 CONTROLS TICK Y POS
     col = cols_x, pch = 15, cex = 9.5,
     xlab = "Value", ylab = "", yaxt = "n",
     xaxt = "n", bty = "n",
     xlim = c(val_min, val_max),  # force exact alignment
     ylim = c(0, 1))              # thin strip

# --- Helper for scientific axis labels ---
sci_labels <- function(vals, digits = 2) {
  labs <- vector("list", length(vals))
  for (i in seq_along(vals)) {
    x <- vals[i]
    if (x == 0) {
      labs[[i]] <- expression(0)
    } else {
      expo <- floor(log10(abs(x)))
      base <- signif(x / 10^expo, digits)
      labs[[i]] <- bquote(.(base) %*% 10^.(expo))
    }
  }
  as.expression(labs)
}
ticks <- c(val_min, (val_min + val_max)/2, val_max)
# draw only ticks + labels, no baseline
axis(1, at = ticks, labels = sci_labels(ticks),
     lwd = 0,       # removes horizontal axis line
     lwd.ticks = 2) # keeps tick marks

# ticks: min, mid, max
# axis(1, at = c(val_min, (val_min + val_max)/2, val_max),
#      labels = round(c(val_min, (val_min + val_max)/2, val_max), 2))




# --- Custom vertical legend bar ---

# val_min <- min(tip.total.changes[tip.total.changes > 0]) 
# val_max <- max(tip.total.changes, na.rm = TRUE)

val_min <- min(tip.total.changes_log) 
val_max <- max(tip.total.changes_log, na.rm = TRUE)

vals_y <- seq(val_min, val_max, length.out = 200)
cols_y <- map_to_color(vals_y, pal_info)

# user controls
bar_x   <- 0.5      # horizontal position of bar
bar_len <- c(val_min, val_max)  # span of values
bar_th  <- 2        # thickness multiplier (cex for pch=15)

par(mar = c(4, 4, 2, 2))
plot(NA, xlim = c(0, 1), ylim = c(val_min, val_max),
     xaxt = "n", yaxt = "n", xlab = "", ylab = NA, bty = "n")

# plot color bar as vertical strip
points(rep(bar_x, length(vals_y)), vals_y,
       col = cols_y, pch = 15, cex = bar_th)

# ticks: min, mid, max
ticks <- c(val_min, (val_min + val_max)/2, val_max)
axis(2, at = ticks, labels = sci_labels(ticks),
     lwd = 0,       # no baseline line
     lwd.ticks = 1, las = 1,  # keep tick marks
     line = -0.9) # controls y position

dev.off()

```


### Useful cols
```{r}
"mediumorchid3"
"darkorange2"
"goldenrod2"
"gold"
"purple3" 
"forestgreen"
"deepskyblue3"
```



## Branch enrichment

###  enrichment level 1
```{r}
enr <- get_enrichment_mask(rates, lower_q = 0.05, upper_q = 0.95, direction = "over", return_type = "binary")
enr.branch <- apply(enr, 1, sum)
enr.boolean <- (enr.branch > 0)*1
hist(enr.branch)

plot_branch_colored_tree(tree, enr.boolean, palette = viridis(15), 
                         title = "all", legend_title = 'Cols')
```

### enrichment level 2

```{r}
#------ Convert to binary
enr.bin <- apply(enr, 1, function(x) paste0(x, collapse = ''))
# enr.bin
unique(enr.bin)
sum(enr.bin > 0)
# enr.boolean

# region_class is sourced from the file
region_class
enr_class <- enrich_next_level(enr, region_class)
enr_class <- enr_class[c("head", "mesosoma", "metasoma", "legs",  "wings")]
head(enr_class)
#head(enr)

# Get enrichment for 5 regions
bin.enr <- apply(enr_class, 1, function(x) paste0(x, collapse = ''))
bin.enr[grepl("^0+$", bin.enr)] <- ""
unique(bin.enr) %>% length()
table(bin.enr)
#sum(enr_class > 0)

plot_branch_colored_tree(tree, enr.boolean, palette = viridis(15), 
                         title = "all", legend_title = 'Cols')


edgelabels(bin.enr, frame = "none", col = "red", cex = 0.6)

```


## Plot tree

```{r}
# ---- Save to EPS ----
#if (!dir.exists("plots")) dir.create("plots")
file_base <- "figures/tree-enrich/tree_enrich"

# postscript(
#   file = paste0(file_base, ".eps"),
#   width = 20,          # width in inches (adjust for column width)
#   height = 9.5,         # height in inches
#   horizontal = FALSE, # portrait orientation
#   onefile = FALSE,    # one figure per file
#   paper = "special"   # use exact width/height
# )


pdf(
  file = paste0(file_base, ".pdf"),
  width = 20, height = 9.5
)

# png(
#   file = paste0(file_base, ".png"),
#   width = 20, height = 9.5, units = "in", res = 300
# )

# ==========================
# Plot tree and color branches
# ==========================
branch_colors <- ifelse(enr.boolean == 1, "#2ca02c", "black")
branch_widths <- ifelse(enr.boolean == 1, 6, 2.0)  # thicker if enriched


plot(tree, edge.color = branch_colors, show.tip.label = TRUE, 
     cex = 0.5, label.offset = 90, edge.width = branch_widths)

# ==========================
# Plot heatmap
# ==========================

lastPP <- get("last_plot.phylo", envir = .PlotPhyloEnv)



# Tip coordinates
tip_y <- lastPP$yy[1:Ntip(tree)]
tip_x <- max(lastPP$xx)



# ---- Square sizes ----
dy <- median(diff(sort(tip_y)))
box_size_y <- dy * 1.2   # adjust multiplier for bigger/smaller squares

# Scaling factor: keep squares true
pin <- par("pin")  # plot dimensions (inches)
xrange <- diff(lastPP$x.lim)
yrange <- diff(lastPP$y.lim)
x_per_inch <- xrange / pin[1]
y_per_inch <- yrange / pin[2]
box_size_x <- box_size_y * (x_per_inch / y_per_inch)

# ---- Gap between tree and heatmap ----
gap <- 0.025 * max(lastPP$xx)

# ---- Define groups of columns ----
# Example: 3 groups 1:2, 3:8, 9:15
groups <- list(1:2, 3:8, 9:10, 11:13, 14:15, 16:17)

# Compute group offsets
gap_size <- box_size_x * 0.4   # adjust for wider/narrower gaps
group_offsets <- rep(0, ncol(tip.total.changes_log))
for (g in seq_along(groups)) {
  if (g > 1) {
    group_offsets[groups[[g]]] <- (g-1) * gap_size
  }
}

# ---- Draw glyphs ----
for (i in 1:Ntip(tree)) {
  values <- as.numeric(tip.total.changes_log[i, ])
  n <- length(values)
  for (j in 1:n) {
    val <- values[j]
    col <- map_to_color(val, pal_info)
    
    # Add group offset to x position
    xleft   <- tip_x + gap + (j-1) * box_size_x + group_offsets[j]
    xright  <- xleft + box_size_x
    ybottom <- tip_y[i] - box_size_y/2
    ytop    <- tip_y[i] + box_size_y/2
    
    rect(xleft, ybottom, xright, ytop,
         col = col, border = "black", lwd = 0.4)
  }
}



# ==========================
# Plot navajo enrichment on branches
# ==========================

edge_coords <- get("last_plot.phylo", envir = .PlotPhyloEnv)

for (i in seq_along(bin.enr)) {
  state <- bin.enr[i]
  if (state == "") next
  
  # edge endpoints
  x0 <- edge_coords$xx[tree$edge[i, 1]]
  y0 <- edge_coords$yy[tree$edge[i, 1]]
  x1 <- edge_coords$xx[tree$edge[i, 2]]
  y1 <- edge_coords$yy[tree$edge[i, 2]]
  
  # edge midpoint (x only), y = child’s y (keeps glyphs aligned)
  xm <- (x0 + x1) / 2
  ym <- y1
  
  # binary string → vector
  bits <- as.integer(strsplit(state, "")[[1]])
  n <- length(bits)
  
  # rectangle dimensions
  # box_w <- max(edge_coords$xx) * 0.01
  # box_h <- max(edge_coords$yy) * 0.01
  box_w <- max(edge_coords$xx) * 0.011
  box_h <- max(edge_coords$yy) * 0.011
  gap   <- box_w * 0
  
  total_w <- n * (box_w + gap)
  x_start <- xm - total_w/2
  
  # draw boxes
  for (j in seq_len(n)) {
    xleft   <- x_start + (j-1) * (box_w + gap)
    xright  <- xleft + box_w
    ybottom <- ym - box_h/2
    ytop    <- ym + box_h/2
    
    rect(xleft, ybottom, xright, ytop,
         col = ifelse(bits[j] == 1, "seagreen3", "white"),
         border = "black", lwd = 1.3)
  }
}

dev.off()
```

